Consistency scores in text data

### 2020-08-03 outputs/documents/paper.Rmd
Update through to the pre-trained transformer language models section

---
output: 
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: false
    toc: false
    fig_caption: true
    latex_engine: pdflatex
    template: templates/svm-latex-ms.tex
bibliography: "references.bib"
header-includes:
  -  \usepackage{hyperref}
biblio-style: apalike
title: "Consistency scores in text data"
thanks: "Thank you to X, Y and Z for helpful comments. A Shiny application for interactive and small-scale examples is available: https://kelichiu.shinyapps.io/Arianna/. An R package for larger scale applications is available: **HERE**. Our code and datasets are available: https://github.com/RohanAlexander/consistency_scores_in_text_datasets. Comments on the `r format(Sys.time(), '%d %B %Y')` version of this paper are welcome at: rohan.alexander@utoronto.ca."
author:
- name: Ke-Li Chiu
  affiliation: University of Toronto
- name: Rohan Alexander
  affiliation: University of Toronto
abstract: "In this paper we introduce a process to clean the text extracted from PDFs using various methods from natural language processing. Our approach compares originally extracted text with the text generated from, or expected by, these methods using earlier text as stimulus. To guide this process, we introduce the notion of a consistency score, which refers to the proportion of text that is unchanged by the model. This is used to monitor changes during the cleaning process, and to compare the messiness of different texts. The methods that we consider are: n-grams, Google's Word2vec, Stanford's GloVe, and OpenAI's GPT-2 and GPT-3. We illustrate our process on text from the Canadian Hansard and introduce both a Shiny application and an R package to make our process easier for others to adopt."
keywords: "text-as-data; natural language processing; quantitative analysis."
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: mathpazo
fontsize: 12pt
# spacing: double
endnote: no
graphics: yes
---

@@

We consider various natural language processing methods. These include n-grams, Google's Word2vec, Stanford's GloVe, and OpenAI's GPT-2 and GPT-3. The n-gram approach involves identifying two, three, or more, words that are commonly found together. Here think of a two-gram involving the word 'good' 'good morning'. At scale these can identify missing or unusual words, and work quickly, but they lack nuance. For instance an equally reasonable two-gram involving the word 'good' is 'good work'. For that reason we consider pre-trained word embedding models. These include Google's Word2vec and Stanford's GloVe, which place each word in a multi-dimensional space such that distance between words can illustrate their relationship. We also consider pre-trained generative models such as OpenAI's GPT-2 and GPT-3. GPT-2 and GPT-3 are pre-trained generative unsupervised language models. They differ in the number of parameters. Generative Pretrained Transformer (GPT)-2 has 1.5 billion parameters and GPT-3 has more than 175 billion parameters. While GPT-2 is more limited, it can be run on smaller machines, whereas GPT-3 cannot and is accessed via an API.

@@

Given the nature of human languages, some combinations of words tend to occur more frequently than others. Think of 'good', which is more often followed by 'morning', than 'duck'. As such, we could consider English text production as a conditional probability, $\mbox{Pr}(w_{k} | w^{k-1}_{1})$, where $k$ is the number of words in a sentence, $w_{k}$ is the predicted word, and $w^{k-1}_{1}$ is the history of the word occurring in a sequence [@brown1992class]. In this way, the generation of some prediction, $w_{k}$, is based on the history, $w^{k-1}_{1}$. This is the underlying principle of all language models. Essentially, a language model is a probability distribution over sequences of words. The goal of statistical language modeling is to estimate probability distributions over different linguistic units — words, sentences, and even documents (@bengio2003neural). However, this is difficult as language is categorical. If we consider each word in a vocabulary as a category, then the dimensionality of language becomes large [@rosenfeld2000two]. The reason that there is scuh a variety of statistical language models is that there are various ways of dealing with this fundamental problem.

@@

The Portable Document Format (PDF) is typically used to render documents in a relatively  fixed way. PDFs contain a lot of information, but often that information is not able to be analysed directly by statistical language models. The information first needs to be extracted from the PDF, often by optical character recognition (OCR), and then cleaned, prepared, and made into some type of dataset. One issue in this process is that OCR is not perfect. Hence there is a need to correct the output before it is analysed. One way to do this is to correct errors one-by-one and independent of context. We all know that 'tbe' is wrong, and should be 'the'. But statistical language models have the advantage of correction that can occur within context. This allows significant improvements, for instance sometimes 'bat' should be 'hat' and sometimes it really should be 'bat' **(Ke-Li add ref)**.

The foundation of an n-gram language model is the conditional probability set-up introduced above. An n-gram model is a probabilistic language model that predicts the next word in a sequence of words **(Ke-Li add ref)**. The $n$ in n-gram referes to the number of words in that sequence. Consider the following excerpt from *Jane Eyre*:

> 'We had been wandering'.

'We' is a uni-gram, 'We had' is a bi-gram, 'We had been' is a tri-gram, and 'We had been wandering' is a 4-gram. Notice that the two tri-grams in this excerpt, 'We had been', and 'had been wandering', overlap. The use of n-gram models enables us to assign probabilities to both the next sequence of words and just the next word. For instance, consider the two sentences: 'We had been wandering', and 'We had been wangling'. The former is likely to be more frequently encountered in a training corpus. Thus, an n-gram would assign a higher probability to the next word being 'wandering' than 'wangling', given the sequence 'We had been'.

@@

$$\mbox{Pr(We,had,been,wandering)} = \mbox{Pr(We)} \times \mbox{P(had|We)} \times \mbox{P(been|We,had)} \\ 
\times \mbox{P(wandering|We,had,been)}.$$

@@

$$\mbox{Pr(We,had,been,wandering)} \approx \mbox{Pr(We)} \times \mbox{P(had|We)} \times \mbox{P(been|had)} \\ 
\times \mbox{P(wandering|been)}.$$

@@

However, n-gram models do not take the linguistic structure of language into account. For instance, @rosenfeld2000two [p. X]  **(Ke-Li add)** discusses language in this context, saying that '...it may as well be a sequence of arbitrary symbols, with no deep structure, intention or thought behind'. The prediction of the next word is based on only a few preceding words and broader context is not taken into account. Hence next-word prediction using n-gram based language models can be limited.

## Word Embedding Models

Turning words into numeric representations is an important step for machine to understand and process words. The most basic way to turn a word into a vector is one-hot encoding. However, with one-hot encoding, the number of dimensions depends on the size of the vocabulary in a given corpus, and can easily grow to extremely large number of dimensions. Word embedding models work by representing words as numerical dense vectors so the dimensions are reduced by representing the words by features (@bengio2003neural). Representing words as dense vectors has a long history, but the idea of a dense embedding space for words was first explored in 2003 by @bengio2003neural, and became the prominent driver for word embedding models. In the work of @bengio2003neural, words are encoded into feature vectors that represents different aspects of the word, and each word is associated with a point in a vector space. The words that are similar will have similar representations in a vector space. The similarity of the words, however, does not depend on their spelling or forms, but by their co-occurrence in given contexts. For example, a "cat" is more similar to a "dog" instead of a "car" because we are more likely to talk about cats and dogs in a same conversation. Therefore, the co-occurrence of "cat" and "dog" has higher probability than the co-occurrence of "cat" and a "car". In other words, the distributed representation is learned by the usage of words. This permits words that are used in similar contexts to result in having similar vector representations and their semantic meanings are captured naturally. As a result, word embedding models are capable of finding the most similar words, calculate the similarities between a pair of words, and find the odd word in a group of words @mikolov2013efficient. The most commonly used word embedding models are Word2Vec and GloVe, which will be described in the following sections.

### Word2Vec

Developed by Google, Word2Vec is a word embedding model that uses shallow neural network. While it is not a deep learning model, it turns words into a vector representations that deep neural networks can understand. It takes a corpus as its input and output a set of feature vectors that represent the words in the given corpus. @mikolov2013efficient proposed two architectures for learning distributed representations of words within the framework of Word2Vec — Continuous Bag-of-Words Model (CBOW) and Continuous Skip-gram Model. Both architectures contains three layers: input layer, projection layer and output layer. The input layer encodes words within a context window to one-hot vectors. Using a projection matrix, the input layer is  projected to a projection layer and the features are. Finally, the output layer is a softmax regression classifier that obtains the posterior distribution of words (@rong2014word2vec). The input and output vectors differ in the two architectures. CBOW uses several surrounding words, or the context, to predict a target word. On the other hand, continuous skip-gram takes a single word as input to predict its neighbouring words. For example, consider the sentence "I think therefore I am", a famous quote by Descartes. The focus word is we want to learn the embedding is set to "therefore", and the context window is set to two which will give us the two immediately preceding words and two immediate proceeding words of the focus word. The input for the CBOW is "I, think, I, am" and the output is "therefore". In contrast, the skip-gram model will take "therefore" as the input and out "I, think, I, am" as the prediction.

Word2Vec is trained on the Google News data set that contains 6 billion tokens and the vocabulary size is restricted to one million most common words (@mikolov2013efficient). The enormous size of data set catches accurate statistical pattern of language and results a comprehensive mapping of word relations to a vector place. The positions of the word vectors reflect their semantic relationships between each other. Therefore, the Word2Vec model performs especially well in word analogy tasks through answering questions such as "$king$ is to $man$ as $queen$ is to __?" @pennington2014glove. The vectorization of words allows the operation of linear algebra for the word vectors in order to get the answer:

>  vector("King") - vector("Man") + vector("Woman") = vector("Queen")

@@

GloVe stands for global vectors for word representation. It is a pre-trained word embedding model that can capture semantic meanings of words developed by Stanford. In the Word2Vec model, the model learn word presentations locally within a context window. On the other hand, GloVe considers the properties of the corpus globally. The benefit of the global approach is that the model can learn about the global statistics of an entire corpus. GloVe can then obtain the co-occurrence probabilities by creating a co-occurrence count matrix for the entire training data set. However, GloVe goes beyond calculating distribution of word by the probability of co-occurrence; it reveals properties of words by the ratio of the probability of co-occurrence of words. For example, consider the two words we want to obtain the word vectors from "ice" and "steam". GloVe reveals their properties by examining their co-occurrence with a variety of probe words, $k$. $K$ can be "solid", which is the property of ice; "gas", which is the property of steam; and "water", which is related to both "ice" and "steam". The ratio of co-occurrence probability is calculated as the following:

> P(k|ice)/P(k|steam)

The ratio between the co-occurrence probability of "ice, solid" pair and "steam, solid" pair is high because the co-occurrence of "ice" and "solid" is higher than the co-occurrence of "steam" and "solid". Reversely, the ratio between the co-occurrence of "ice, gas" pair and "steam, gas" pair is low. On the other hand, the ratio of co-occurrence probability of "ice, water" pair and "steam, water" pair are close to 1 because water is relevant to both "ice" and "steam". Probability alone does not distinguish well relevant words from irrelevant words. For this reason, the researcher of GloVe argued that "the appropriate starting point for word vector learning should be with ratios of co-occurrence probabilities rather than the probabilities themselves (@pennington2014glove)."

@@

In recent years, the use of pre-trained language models has played a significant role in advancing natural language processing (NLP) tasks such as reading comprehension, text generation, and even creative writing (@brown2020language). Pre-trained models remove the need for researchers to prepare training datasets or allot computational resources to the stage of pre-training. Because of this, there has been substantial development in NLP research. OpenAI and Google are two companies that are considered pioneers in artificial intelligence and they both develop pre-trained models that are considered state-of-art; we will discuss the GPT models from OpenAI and the BERT model from Google.

>>>>

---
output: 
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: false
    toc: false
    fig_caption: true
    latex_engine: pdflatex
    template: templates/svm-latex-ms.tex
bibliography: "references.bib"
header-includes:
  -  \usepackage{hyperref}
biblio-style: apalike
title: "Consistency scores in text data"
thanks: "Thank you to **X, Y and Z** for helpful comments. A Shiny application for interactive and small-scale examples is available: https://kelichiu.shinyapps.io/Arianna/. An R package for larger scale applications is available: **HERE**. Our code and datasets are available: https://github.com/RohanAlexander/consistency_scores_in_text_datasets. Comments on the `r format(Sys.time(), '%d %B %Y')` version of this paper are welcome at: rohan.alexander@utoronto.ca."
author:
- name: Ke-Li Chiu
  affiliation: University of Toronto
- name: Rohan Alexander
  affiliation: University of Toronto
abstract: "In this paper we introduce a process to clean the text extracted from PDFs using various methods from natural language processing. Our approach compares originally extracted text with the text generated from, or expected by, these methods using earlier text as stimulus. To guide this process, we introduce the notion of a consistency score, which refers to the proportion of text that is unchanged by the model. This is used to monitor changes during the cleaning process, and to compare the messiness of different texts. The methods that we consider are: n-grams, word2vec, GloVe, and the GPT models. We illustrate our process on text from the Canadian Hansard and introduce both a Shiny application and an R package to make our process easier for others to adopt."
keywords: "text-as-data; natural language processing; quantitative analysis."
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: mathpazo
fontsize: 12pt
# spacing: double
endnote: no
graphics: yes
---

@@

We consider various natural language processing methods. These include n-grams, word2vec, GloVe, and the GPT models. The n-gram approach involves identifying two, three, or more, words that are commonly found together. Here think of a two-gram involving the word 'good' 'good morning'. At scale these can identify missing or unusual words, and work quickly, but they lack nuance. For instance, an equally reasonable two-gram involving the word 'good' is 'good work'. For that reason, we consider pre-trained word embedding models. These include word2vec and GloVe, which place each word in a multi-dimensional space such that distance between words can illustrate their relationship. We also consider pre-trained generative models such as GPT-2 and GPT-3. GPT-2 and GPT-3 are pre-trained generative unsupervised language models. They differ in the number of parameters. GPT-2 has 1.5 billion parameters and GPT-3 has more than 175 billion parameters. While GPT-2 is more limited, it can be run on smaller machines, whereas GPT-3 cannot and is accessed via an API.

@@

Given the nature of human languages, some combinations of words tend to occur more frequently than others. Think of 'good', which is more often followed by 'morning', than 'duck'. As such, we could consider English text production as a conditional probability, $\mbox{Pr}(w_{k} | w^{k-1}_{1})$, where $k$ is the number of words in a sentence, $w_{k}$ is the predicted word, and $w^{k-1}_{1}$ is the history of the word occurring in a sequence [@brown1992class]. In this way, the generation of some prediction, $w_{k}$, is based on the history, $w^{k-1}_{1}$. This is the underlying principle of all language models. Essentially, a language model is a probability distribution over sequences of words. The goal of statistical language modeling is to estimate probability distributions over different linguistic units — words, sentences, and even documents [@bengio2003neural]. However, this is difficult as language is categorical. If we consider each word in a vocabulary as a category, then the dimensionality of language becomes large [@rosenfeld2000two]. The reason that there is such a variety of statistical language models is that there are various ways of dealing with this fundamental problem.

@@

The Portable Document Format (PDF) is typically used to render documents in a relatively fixed way. PDFs contain a lot of information, but often that information is not able to be analysed directly by statistical language models. The information first needs to be extracted from the PDF, often by optical character recognition (OCR), and then cleaned, prepared, and made into some type of dataset. One issue in this process is that OCR is not perfect. Hence there is a need to correct the output before it is analysed. One way to do this is to correct errors one-by-one and independent of context. We all know that 'tbe' is wrong and should be 'the'. But statistical language models have the advantage of correction that can occur within context. This allows significant improvements, for instance sometimes 'bat' should be 'hat' and sometimes it really should be 'bat' **(Ke-Li add ref)**.

The foundation of an n-gram language model is the conditional probability set-up introduced above. An n-gram model is a probabilistic language model that predicts the next word in a sequence of words **(Ke-Li add ref)**. The $n$ in n-gram refers to the number of words in that sequence. Consider the following excerpt from *Jane Eyre*: 'We had been wandering'. 'We' is a uni-gram, 'We had' is a bi-gram, 'We had been' is a tri-gram, and 'We had been wandering' is a 4-gram. Notice that the two tri-grams in this excerpt, 'We had been', and 'had been wandering', overlap. The use of n-gram models enables us to assign probabilities to both the next sequence of words and just the next word. For instance, consider the two sentences: 'We had been wandering', and 'We had been wangling'. The former is likely to be more frequently encountered in a training corpus. Thus, an n-gram would assign a higher probability to the next word being 'wandering' than 'wangling', given the sequence 'We had been'.

@@

$$\mbox{Pr(We,had,been,wandering)} = \mbox{Pr(We)} \times \mbox{Pr(had|We)} \times \mbox{Pr(been|We,had)} \\ 
\times \mbox{Pr(wandering|We,had,been)}.$$

@@

$$\mbox{Pr(We,had,been,wandering)} \approx \mbox{Pr(We)} \times \mbox{Pr(had|We)} \times \mbox{Pr(been|had)} \\ 
\times \mbox{Pr(wandering|been)}.$$

@@

However, n-gram models do not take the linguistic structure of language into account. For instance, @rosenfeld2000two [p. X]  **(Ke-Li add page please)** discusses language in this context, saying that '...it may as well be a sequence of arbitrary symbols, with no deep structure, intention or thought behind'. The prediction of the next word is based on only a few preceding words and broader context is not taken into account. Hence next-word prediction using n-gram based language models can be limited.

**(Ke-Li add brief write-up of implementation in R please. Probably quanteda unless there's a better option.)**

## Word embedding models

Although, from an implementation perspective, there is an inherent numerical representation of language in n-gram models, it is not overly tractable. The numeric representation typically relies on each term being represented in a vector, which can quickly result in quite high-dimensional objects. In contrast, word embedding models represent terms by vectors of real numbers where each vector is a 'feature', hence the final object has considerably lower dimensionality [@bengio2003neural].

This approach has a long history, but the implementation of @bengio2003neural has become the foundation for much subsequent work. Here, terms are encoded into feature vectors that represent different aspects of the term, and so each term is represented by a position in the vector space. Terms that are used in similar ways will have similar representations in the vector space. This will be due to their co-occurrence in usage. For example, 'cat' will be closer to 'dog' than to 'car', because we are more likely to talk about cats and dogs, than cats and cars. See how they might be close when considered along a 'pets' dimension, but far when considered along a 'related-to-tigers' dimension.

Terms that are often used in similar contexts will have similar vector representations. Hence, word embedding models can identify similar terms, calculate the similarity between a pair of terms, and find the odd term in a group of terms [@mikolov2013efficient]. There are a variety of ways to achieve this dimensionality reduction, including word2vec and GloVe.

### Word2vec

Word2vec is a word embedding model that uses a neural network to learn word associations [@mikolov2013efficient] [@mikolov2013distributed]. It takes a corpus as an input and outputs a set of feature vectors that allow the terms in that corpus to be represented. @mikolov2013efficient proposed two architectures for determining this distributed representation of terms: continuous bag-of-words (CBOW), and continuous skip-gram. Both have an input layer, a projection layer, and an output layer. But CBOW uses several surrounding words, or the context, to predict a target word. On the other hand, continuous skip-gram takes a single word as input to predict its neighbouring words.

To illustrate the difference, consider the quote from Descartes, 'I think therefore I am'. The focus word for which we want to learn the embedding is 'therefore', and the context window is set to two. This means that we consider the two words immediately preceding and proceeding 'therefore'. The CBOW input is 'I, think, I, am', and the output is 'therefore'. In contrast, the continuous skip-gram input is 'therefore', while 'I, think, I, am' is the output. The input layer encodes terms within a context window to one-hot vectors. Using a projection matrix, the input layer is  projected to a projection layer and the features are. **(Ke-Li: Please fix this sentence.)** Finally, the output layer is a softmax regression classifier that obtains the posterior distribution of words (@rong2014word2vec). It is the input and output vectors that differ between the two architectures.

Word2vec is trained on the Google News dataset **(Ke-Li add cite to the dataset please.)**. This contains six billion tokens and the vocabulary size is restricted to the one million most common words (@mikolov2013efficient). As this is a large dataset, it provides an accurate statistical pattern of language, as it is used in this dataset, and a comprehensive mapping between terms and vector positions.

To the extent that the relative positions of terms reflects their semantic relationship, word2vec will perform well for word analogies. However, terms that have context-specific meaning may not perform as well.

**(Ke-Li add brief write-up of implementation in R please.)**

@@

The global vectors (GloVe) model is a pre-trained word embedding model that emphasises the semantic meaning of words [@pennington2014glove]. In the word2vec model, word representations are learnt locally within the context window. On the other hand, GloVe considers the properties of the corpus globally. As the model considers the context of an entire corpus the co-occurrence probabilities are based on counts for the entire training dataset.

GloVe additionally considers the relative probability of word co-occurrence. For instance, consider the two words 'ice' and 'steam'. GloVe considers their co-occurrence with a variety of probe words, $k$. For instance $k$ could be 'solid', which is a property of ice; 'gas', which is a property of steam; and 'water', which is related to both.The ratio of co-occurrence probability is:

$$\frac{\mbox{Pr}(k|\mbox{ice})}{\mbox{Pr}(k|\mbox{steam})}.$$

The ratio between the co-occurrence probability of the 'ice, solid' pair and the 'steam, solid' pair is high because the co-occurrence of 'ice' and 'solid' is higher than the co-occurrence of 'steam' and 'solid'. Conversely, the ratio between the co-occurrence of the 'ice, gas' pair and the 'steam, gas' pair is low. On the other hand, the ratio of the co-occurrence probability of the 'ice, water' pair and the 'steam, water' pair us close to one because water is relevant to both 'ice' and 'steam'.

**(Ke-Li add brief paragraph about negatives of this approach.)**

**(Ke-Li add brief write-up of implementation in R please.)**

@@

In recent years, the use of pre-trained language models has played a significant role in advancing natural language processing (NLP) tasks such as reading comprehension, text generation, and even creative writing [@brown2020language]. Pre-trained models remove the need for researchers to prepare training datasets or allot computational resources to the stage of pre-training. Because of this, there has been substantial development in NLP research. In this section we discuss the GPT models from OpenAI and the BERT model from Google.
